 

### 2021-05-24 vertical-thermal-review.Rmd
editing biophysical section, knit in latest leaf T fig

**The biophysical environment across the vertical gradient from the forest floor to the top of the canopy is in large part determined by the structure of the forest itself.** 
Here, we define the biophysical environment to include the physical structure of the vegetation and associated physical conditions.
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`, Schimel et al. 2007].

### Foliage density
**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and in turn is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
The distribution of this foliage is structured in correspondence with the distribtuion of tree sizes within a forest.  
Leaf area density.... *description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references)*
Vertical profiles in leaf area density vary across forests, being... *summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.

@@

@@

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density. This gradient is most pronounced under dense canopies, including broad-leaf and mixed forests, than open canopy forests such as needle-leaf forests [Fig. `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. Sunflecks, offer another light source for shaded forest regions, caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, and in closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019] wind speeds increase with height. With daily maximum wind speeds averaging up to **#** m s^-1^ higher at the top of the NEON vertical profile than at the bottom

### Air temperature 
**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies. Therefore, dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. (*This sentence is too complex-- hard to understand*)

Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al. 2017; @mcgregor_tree_nodate]. Whereas in colder conditions, dense-canopy understories experience higher minimum temperatures [$T_{min}$] relative to open-canopy understories and nearby clearings where greater radiative heat loss results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al. 2021*; @nakamuraForestsTheirCanopies2017].

### Humidity and evaporative demand
**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a]. $RH$, in combination with temperature, determines vapor pressure deficit [$VPD$]--the driving force of water loss from a leaf-- tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].

@@

With altitude, the strength of buffering decreases, moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for eg. in riparian forests. [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]

>>>>

**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs. `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest itself.** 
In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S. National Ecological Observatory Network [NEON; `r fig_NEON_vertical`; Appendix `r appendix_NEON_methods`, Schimel et al. 2007].
While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment cn be more closely linked to the distance from the outer canopy than to height (Parker 1995).

@@

### Foliage density
**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and in turn is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest.
Leaf area density.... *[description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references).]*
Vertical profiles in leaf area density vary across forests, being... *summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
In this review, we focus on growing season conditions unless otherwise noted.

@@

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the shape of profile modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`, @koike_leaf_2001].** 
This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests, than open canopy forests such as needle-leaf forests [Fig. `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases is solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

### Wind speeds 
**Wind speeds are also higher at the top of the canopy, owing to... [@juckerCanopyStructureTopography2018].**
Across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites, wind speeds increase with height. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

### Air temperature 
**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al. 2017; **double check these refs--at least one didn't belong**].

### Humidity and evaporative demand
**Relative humidity, $RH$, is generally highest in the understory and decreases with height across the vertical profile [Fig. `r fig_NEON_vertical`; e.g., @mcgregor_tree_nodate].**
Along with temperature, dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a]. 
$RH$, in combination with temperature, determines vapor pressure deficit [$VPD$] -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].

@@

**The buffering patterns described above vary geographically and across forests types.**
The strength of buffering decreases with altitude (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) and moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests. [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
